Mainstreaming Race Science

### 2023-10-27 paper/paper.qmd
crossrefs

We identified 16 researchers who had received funding from Pioneer; 14 of these researchers had profiles in the Web of Science (WoS) author search, allowing us identify 13 WoS-indexed journals that had published 6 or more of these authors. See Table @tbl-researchers and Figure @fig:wos.

@@

![Journals publishing 6 or more Pioneer-funded researchers, Web of Science author search results. (\#fig:wos)](img/wos_results.png){width="4.76in" height="2.6in"}

Figure @fig:wos shows that, while MQ is among the "Pioneer-publishing" journals, a number of mainstream journals are more prominent: *Personality and Individual Differences* (PID), *Intelligence* (Int), *Behavior Genetics* (BG), and *Psychological Reports* (PR).  In addition, only psychologist Richard Lynn appears to have published heavily in MQ.  Lynn became an assistant editor of MQ in 1979 (vol. XX, Nos. 1 & 2) and is listed as editor-in-chief on MQ's current website as of 2023-07-21.  He has also been president of Pioneer since the death of psychologist J. Phillippe Rushton in 2012 [@BeirichPioneerFundAssets2013].  By contrast, a number of Pioneer-funded researchers have published in PID, Int, and to a lesser degree BG: Bouchard, Eysenck, Jensen, Rushton, Vernon, and also Lynn.

@@

*Intelligence* (Int) was founded in 1977, with psychologist Douglas Detterman as editor-in-chief from the founding until 2016.  Lloyd Humphreys was on the editorial board starting from 1977; by 1990 he had been joined by Jensen and Philip Vernon.  Richard Lynn joined the editorial board sometime between 1998 and 2002.  (Archive copies of the Int editorial board page are not available from the journal's website from 1999 through 2001.)  Int has been criticized for including Lynn and Gerhard Meisenberg — who was editor-in-chief of MQ in 2015-18 — on its editorial board until 2018 [@SainiSuperiorReturnRace2019}.  Int is published by Elsevier.

# Topic model analysis identifies race science discourse {.unnumbered]

@@

![Silge plots [@SilgeTopicModeling2017] and smoothed time series for three focal topics.  Top row: top 15 terms by $\beta$ (term-topic distribution) for each topic.  Bottom row: count of articles associated with each topic, by journal and year.  Article counts use a threshold approach, with $\gamma$ (topic-document distribution) greater than 0.5.  Thin lines give annual values, thick lines give 5-year running averages.  Because *Behavior Genetics* and *Behavioral and Brain Sciences* are not prominent in any panel, neither is given a direct label. (\#fig:focal)](img/focal_topics.png){width="4.76in" height="3.57in"}

Figure @fig:focal and this discussion focuses on three topics identified in one of the 24 fitted topic models; see figures @fig:gamma-lg, @fig:gamma-md, @fig:gamma-sm and supplemental visualizations S2-S4 for all topics from all models and S5-S7 for selected topics from all models.  In the focal model, topic 07 is strongly associated with MQ:  MQ published dozens of articles in this topic each year, and no other journal ever published more than a handful.  The "Silge plot," showing the top 15 terms in the topic [@SilgeTopicModeling2017], contains racial terms (`races`, `whites`, `negroes` and potentially `europe`, `africa`, `india`, and `japan`) as well as `book`, likely reflecting the fact that MQ published a number of book reviews, while the psychology journals either did not or these were not available for the corpus.

@@

Instead, our results suggest that other subdisciplines of psychology have provided the primary venue for mainstreaming race science.  Over the past few decades, many professional organizations in genetics and anthropology have made formal statements rejecting race as a biologically meaningful concept [@anthropology1996aapa; @@american2018ashg, board1998aaa, committee2020asa; FuentesAAPAStatementRace2019; nhgri2023; rotimi2022ashg; @wynshawboris2020ashg], undermining the core assumption of race science.  The American Psychological Association (APA) --- whose members are predominantly clinical psychologists --- and the Federation of Associations in Behavioral & Brain Sciences (FABBS) --- an organization whose purpose is to provide policy recommendations --- have openly rejected race as biologically meaningful [@american2021apology; @baron2022fabbs].  But research-focused psychological organizations like the Association for Psychological Science (APS) and Psychonomic Society (PS) stop short of rejecting race as biologically meaningful when they denounce racism [@aps2021statement; @board2020psychonomic].  Psychology as a discipline maintains space for several mythological race science narratives, such as allegations of a taboo against race and intelligence research [@jackson2021mythical], claims that scientists engaging in or calling for any kind of anti-racism in the field is an ideological corruption of dispassionate and value-neutral science [@roberts_2022], or arguments that holding race science to the same evidentiary standards as other psychological research is a violation of academic freedom [@herbert2023academic].  And historically several race scientists have secured important gatekeeping positions within the scientific community, such as on the editorial boards of Int and PID.

@@

We are not identifying any particular scientist as a racist, scientific racist, or race scientist based on the topic model results alone.  Such claims require an analysis of documentary evidence that goes beyond the scope of the current study [@JacksonJr.ArthurJensenEvolutionary2022].  We do assume that our readers, like us, regard white supremacy and scientific racism as morally odious and beyond the scope of reasonable debate [@SchroederLimitsDemocratizingScience2022].  However, scientific research can be appropriated to promote scientific racism — and thus count as race science — even when this is contrary to the intentions of the original researchers themselves [@TaberyWhyStudyingGenetics2015; @GillbornSoftlySoftlyGenetics2016; CarlsonQuantifyingContextualizingImpact2020; @HennWhyDNANo2021].  Panofsky et al. show how the term "human biodiversity," originally developed for anti-racist purposes by biological anthropologist Jonathan Marks, has been inverted by white supremacists and is now a dogwhistle for biological racial hierarchy [@PanofskyHowWhiteNationalists2021].

Research on the genetics of intelligence has been socially harmful, not just in the somewhat abstract sense of promoting racial stigma, but in the concrete sense of being used to rationalize mass shootings and other acts of racial violence [@PronczukRacistResearcherExposed2022; @MeyerWrestlingSocialBehavioral2023].  While scientists do not fully control the downstream social effects of their research, like other citizens scientists are responsible for mitigating reasonably foreseeable harms that result from their actions [@BlockIQHeritabilityInequality1974a; @DouglasSciencePolicyValuefree2009; KitcherArgumentFreeInquiry1997; @KouranyShouldKnowledgeBe2016].  These actions include not only individual decisions to research this topic or that, but also collective decisions about who is awarded positions of power or influence.  Given the way research on intelligence and behavior genetics has been used historically, these fields may be especially susceptible to appropriation by scientific racists.  This only strengthens the obligations of researchers in these areas to prevent their research from causing harm [@CarlsonQuantifyingContextualizingImpact2020; @OgbunugaforDNABasketballBirthday2022].

@@

# References and Notes {.unnumbered}

\bibliographystyle{Science}
\bibliography{references.bib, refs-manual.bib}

| 
| Thanks to Emily Merchant and John Jackson for some initial discussions that helped clarify the scope of the project and identify some key resources related to the Pioneer Fund and *Mankind Quarterly*.  Thanks to Anthony Sainez for help retrieving the *Mankind Quarterly* articles.  Thanks to Derek Devnich and James Dooley for their work attempting to secure the APA-published articles.  Thanks to Kevin Bird, John Jackson, and Manuela Fernández Pinto for comments on a draft of this paper. 
| **Funding:** EJCL's work on this project was partially supported by UC Merced.  DJH's work on this project was not supported by any specific funding. 
| **Author contributions:** DJH (conceptualization, data curation, formal analysis, investigation, methodology, project administration, software, validation, visualization, writing-original draft, writing-review & editing); EJCL (conceptualization, methodology, software, validation, visualization, writing-original draft, writing-review & editing)
| **Competing interests:** The authors have no competing interests to declare. 
| **Data and materials availability:** Web of Science search results and article fulltext are not publicly available due to intellectual property restrictions.  Preprocessed analysis data (document-term counts and article metadata) and analysis scripts are available at *[doi]*.

@@

| S1. Materials and methods
| Figures S2-S7
| Data S8, S9

@@

![Corpus assembly. (\#fig:corpus)](img/dataset.png){width="8in" height="6.1in"}

Corpus assembly is summarized in figure @fig:corpus.  The corpus was assembled in two parts, one for *Mankind Quarterly* and the other for mainstream journals.

@@

After XML/PDF versions were retrieved, text was extracted using either the `xml2` package [@WickhamXml2ParseXML2023] or `pdftools`.  Figure @fig:counts and table @tbl-counts show the number of fulltext documents included in the corpus.  All together, the corpus includes 34,896 documents from 6 journals.

![Count of documents in the corpus, by journal and year. (\#fig:counts)](img/02_count.png){width="4.76in" height="3.6in"}

Table: (\#tbl-counts) Document counts, by journal, and years included in the corpus.

@@

To fit topic models, we followed the approach proposed by Rohe and Zeng [@RoheVintageFactorAnalysis2020], which uses varimax-rotated partial principal components instead of the variational inference methods used by standard topic model packages such as `stm` [@RobertsStmPackageStructural2019}.  This novel approach was implemented in the R package `tmfast` [HicksTmfastFitsTopic2023].  We conducted a simulation study of `tmfast`, which found that it was significantly faster and only slightly less accurate at reconstructing known word-topic and topic-document distributions, compared to `stm` [@HicksTmfastFitsTopic2023].

@@

After fitting topic models, our first research question was whether we could identify distinctive "race science discourse" topics.  Figures @fig:gamma-lg, @fig:gamma-md, @fig:gamma-sm show the gamma (topic-document) distributions for each value of $k$ for the three vocabularies.  In these figures, each panel corresponds to a single journal-$k$ combination, each row of cells is a single document in that journal, and each column of cells is a single topic in that particular model.  Topics correspond within columns of panels, but not within rows; for example, topic 04 for $k=10$ does not necessarily correspond to topic 04 for $k=20$.  Color intensity indicates the value of gamma for that particular topic-document combination.

![Fitted topic models: Gamma (topic-document) distributions, large vocabulary. (\#fig:gamma-lg)](img/05_lg_gamma.png){width="6in" height="4.8in"]

![Fitted topic models: Gamma (topic-document) distributions, medium vocabulary. (\#fig:gamma-md)](img/05_md_gamma.png){width="6in" height="4.8in"}

![Fitted topic models: Gamma (topic-document) distributions, small vocabulary. (\#fig:gamma-sm)](img/05_sm_gamma.png){width="6in" height="4.8in"}

@@

<!-- 
\clearpage

# References and Notes {.unnumbered}

\bibliographystyle{Science}
\bibliography{references.bib, refs-manual.bib}

 -->

>>>>

We identified 16 researchers who had received funding from Pioneer; 14 of these researchers had profiles in the Web of Science (WoS) author search, allowing us identify 13 WoS-indexed journals that had published 6 or more of these authors. See @tbl-researchers and @fig-wos.

@@

![Journals publishing 6 or more Pioneer-funded researchers, Web of Science author search results.](img/wos_results.png){#fig-wos width="4.76in" height="2.6in"}

@fig-wos shows that, while MQ is among the "Pioneer-publishing" journals, a number of mainstream journals are more prominent: *Personality and Individual Differences* (PID), *Intelligence* (Int), *Behavior Genetics* (BG), and *Psychological Reports* (PR).  In addition, only psychologist Richard Lynn appears to have published heavily in MQ.  Lynn became an assistant editor of MQ in 1979 (vol. XX, Nos. 1 & 2) and is listed as editor-in-chief on MQ's current website as of 2023-07-21.  He has also been president of Pioneer since the death of psychologist J. Phillippe Rushton in 2012 [@BeirichPioneerFundAssets2013].  By contrast, a number of Pioneer-funded researchers have published in PID, Int, and to a lesser degree BG: Bouchard, Eysenck, Jensen, Rushton, Vernon, and also Lynn.

@@

*Intelligence* (Int) was founded in 1977, with psychologist Douglas Detterman as editor-in-chief from the founding until 2016.  Lloyd Humphreys was on the editorial board starting from 1977; by 1990 he had been joined by Jensen and Philip Vernon.  Richard Lynn joined the editorial board sometime between 1998 and 2002.  (Archive copies of the Int editorial board page are not available from the journal's website from 1999 through 2001.)  Int has been criticized for including Lynn and Gerhard Meisenberg — who was editor-in-chief of MQ in 2015-18 — on its editorial board until 2018 [@SainiSuperiorReturnRace2019].  Int is published by Elsevier.

# Topic model analysis identifies race science discourse {.unnumbered}

@@

![Silge plots [@SilgeTopicModeling2017] and smoothed time series for three focal topics.  Top row: top 15 terms by $\beta$ (term-topic distribution) for each topic.  Bottom row: count of articles associated with each topic, by journal and year.  Article counts use a threshold approach, with $\gamma$ (topic-document distribution) greater than 0.5.  Thin lines give annual values, thick lines give 5-year running averages.  Because *Behavior Genetics* and *Behavioral and Brain Sciences* are not prominent in any panel, neither is given a direct label.](img/focal_topics.png){#fig-focal width="4.76in" height="3.57in"}

@fig-focal and this discussion focuses on three topics identified in one of the 24 fitted topic models; see figures @fig-gamma-lg, @fig-gamma-md, @fig-gamma-sm and supplemental visualizations S2-S4 for all topics from all models and S5-S7 for selected topics from all models.  In the focal model, topic 07 is strongly associated with MQ:  MQ published dozens of articles in this topic each year, and no other journal ever published more than a handful.  The "Silge plot," showing the top 15 terms in the topic [@SilgeTopicModeling2017], contains racial terms (`races`, `whites`, `negroes` and potentially `europe`, `africa`, `india`, and `japan`) as well as `book`, likely reflecting the fact that MQ published a number of book reviews, while the psychology journals either did not or these were not available for the corpus.

@@

Instead, our results suggest that other subdisciplines of psychology have provided the primary venue for mainstreaming race science.  Over the past few decades, many professional organizations in genetics and anthropology have made formal statements rejecting race as a biologically meaningful concept [@anthropology1996aapa; @american2018ashg; @board1998aaa; @committee2020asa; @FuentesAAPAStatementRace2019; @nhgri2023; @rotimi2022ashg; @wynshawboris2020ashg], undermining the core assumption of race science.  The American Psychological Association (APA) --- whose members are predominantly clinical psychologists --- and the Federation of Associations in Behavioral & Brain Sciences (FABBS) --- an organization whose purpose is to provide policy recommendations --- have openly rejected race as biologically meaningful [@american2021apology; @baron2022fabbs].  But research-focused psychological organizations like the Association for Psychological Science (APS) and Psychonomic Society (PS) stop short of rejecting race as biologically meaningful when they denounce racism [@aps2021statement; @board2020psychonomic].  Psychology as a discipline maintains space for several mythological race science narratives, such as allegations of a taboo against race and intelligence research [@jackson2021mythical], claims that scientists engaging in or calling for any kind of anti-racism in the field is an ideological corruption of dispassionate and value-neutral science [@roberts_2022], or arguments that holding race science to the same evidentiary standards as other psychological research is a violation of academic freedom [@herbert2023academic].  And historically several race scientists have secured important gatekeeping positions within the scientific community, such as on the editorial boards of Int and PID.

@@

We are not identifying any particular scientist as a racist, scientific racist, or race scientist based on the topic model results alone.  Such claims require an analysis of documentary evidence that goes beyond the scope of the current study [@JacksonJr.ArthurJensenEvolutionary2022].  We do assume that our readers, like us, regard white supremacy and scientific racism as morally odious and beyond the scope of reasonable debate [@SchroederLimitsDemocratizingScience2022].  However, scientific research can be appropriated to promote scientific racism — and thus count as race science — even when this is contrary to the intentions of the original researchers themselves [@TaberyWhyStudyingGenetics2015; @GillbornSoftlySoftlyGenetics2016; @CarlsonQuantifyingContextualizingImpact2020; @HennWhyDNANo2021].  Panofsky et al. show how the term "human biodiversity," originally developed for anti-racist purposes by biological anthropologist Jonathan Marks, has been inverted by white supremacists and is now a dogwhistle for biological racial hierarchy [@PanofskyHowWhiteNationalists2021].

Research on the genetics of intelligence has been socially harmful, not just in the somewhat abstract sense of promoting racial stigma, but in the concrete sense of being used to rationalize mass shootings and other acts of racial violence [@PronczukRacistResearcherExposed2022; @MeyerWrestlingSocialBehavioral2023].  While scientists do not fully control the downstream social effects of their research, like other citizens scientists are responsible for mitigating reasonably foreseeable harms that result from their actions [@BlockIQHeritabilityInequality1974a; @DouglasSciencePolicyValuefree2009; @KitcherArgumentFreeInquiry1997; @KouranyShouldKnowledgeBe2016].  These actions include not only individual decisions to research this topic or that, but also collective decisions about who is awarded positions of power or influence.  Given the way research on intelligence and behavior genetics has been used historically, these fields may be especially susceptible to appropriation by scientific racists.  This only strengthens the obligations of researchers in these areas to prevent their research from causing harm [@CarlsonQuantifyingContextualizingImpact2020; @OgbunugaforDNABasketballBirthday2022].

@@

@@

| 
| S1. Materials and methods
| Figures S2-S7
| Data S8, S9

@@

![Corpus assembly.](img/dataset.png){#fig-corpus width="8in" height="6.1in"}

Corpus assembly is summarized in @fig-corpus.  The corpus was assembled in two parts, one for *Mankind Quarterly* and the other for mainstream journals.

@@

After XML/PDF versions were retrieved, text was extracted using either the `xml2` package [@WickhamXml2ParseXML2023] or `pdftools`.  @fig-counts and table @tbl-counts show the number of fulltext documents included in the corpus.  All together, the corpus includes 34,896 documents from 6 journals.

![Count of documents in the corpus, by journal and year.](img/02_count.png){#fig-counts width="4.76in" height="3.6in"}

Table: Document counts, by journal, and years included in the corpus. {#tbl-counts}

@@

To fit topic models, we followed the approach proposed by Rohe and Zeng [@RoheVintageFactorAnalysis2020], which uses varimax-rotated partial principal components instead of the variational inference methods used by standard topic model packages such as `stm` [@RobertsStmPackageStructural2019].  This novel approach was implemented in the R package `tmfast` [@HicksTmfastFitsTopic2023].  We conducted a simulation study of `tmfast`, which found that it was significantly faster and only slightly less accurate at reconstructing known word-topic and topic-document distributions, compared to `stm` [@HicksTmfastFitsTopic2023].

@@

After fitting topic models, our first research question was whether we could identify distinctive "race science discourse" topics.  [@fig-gamma-lg; @fig-gamma-md; and @fig-gamma-sm] show the gamma (topic-document) distributions for each value of $k$ for the three vocabularies.  In these figures, each panel corresponds to a single journal-$k$ combination, each row of cells is a single document in that journal, and each column of cells is a single topic in that particular model.  Topics correspond within columns of panels, but not within rows; for example, topic 04 for $k=10$ does not necessarily correspond to topic 04 for $k=20$.  Color intensity indicates the value of gamma for that particular topic-document combination.

![Fitted topic models: Gamma (topic-document) distributions, large vocabulary.](img/05_lg_gamma.png){#fig-gamma-lg width="6in" height="4.8in"}

![Fitted topic models: Gamma (topic-document) distributions, medium vocabulary.](img/05_md_gamma.png){#fig-gamma-md width="6in" height="4.8in"}

![Fitted topic models: Gamma (topic-document) distributions, small vocabulary.](img/05_sm_gamma.png){#fig-gamma-sm width="6in" height="4.8in"}

@@

# References
